An Introduction to Rocker: Docker Containers for R

### 2017-07-14 manuscript.Rmd
wip

All Rocker images are based on the Debian Linux Distribution.
The Debian platform provides a small base image, the well-known `apt`
package management system and rich ecosystem of software libraries,
making it the base image of choice for Docker images, including many
of the "official" images maintained by Docker's own development team.
The Debian platform is also perhaps the best supported Linux platform
within the R community, including an active `r-sig-debian` listserve.
Unfortunately, the relatively long period between stable Debian releases
(roughly two years recently) means that software in the `debian:stable`
releases can lag significantly behind current releases of popular
software, including R.  More recent versions of packages can be found in
the pre-release distribution, `debian:testing`, while the very latest
binary builds can be found on `debian:unstable`.  The Rocker Project
can be largely divided into two stacks which address different needs,
defined by which Debian distribution they are based on.  The first stack
is based on `debian:testing`, the second, more recently introduced stack
is based only on `debian` stable releases.  These stacks have different
aims and thus provide different images, as shown in Tables 1 & 2.

@@

The `debian:testing` stack aims to makes the most efficient use of
upstream builds -- the precompiled `.deb` binaries provided by the
Debian repositories.  It is both quicker and easier to install software
from binaries, since the package manager (`apt`) manages the necessary
(binary) dependencies and bypasses the time-consuming process of compiling
from source.  Basing this stack on `debian:testing` means that much more
recent versions of commonly-used libraries and compilers are available
as binaries than would be found in a Debian stable release.  In order
to provide access to the most recent avialble binaries, this stack uses
apt-pinning [@apt-pinning] to allow the `apt` package manager to also
install binaries from `debian:unstable`, which represents the most recent,
bleeding edge of packages built for Debian when necessary.  For instance,
the `r-base` image provided by Rocker installs the most recent version of
R as a binary from Debian `unstable`.  Similarly, recent versions of many
popular R packages can also be installed through the package manager,
e.g. `apt-get install r-cran-xml`.  This can be particularly helpful
for packages with external system dependencies (such as `libxml2-dev`
in this example) which cannot be installed from the R console as they 
are system dependencies rather than R packages installed from within R.

As the names `testing` and `unstable` imply, this approach is not
without challenges.  The particular version of any given package can
change as packages move from `unstable` into `testing`. New versions
are sent to `unstable` during the normal course of Debian development.
This can occassionally break an previously working installation command
in a Dockerfile until the maintainer redirects the package manager to
install a the package from the `unstable` sources that could previously
be installed from `testing`, or vice versa (using the `-t` option in
`apt`, see Box 1 on apt-pinning.)

@@

1. These Dockerfiles are easy to develop and extend because almost all software can be installed through the package manager
2. These Dockerfiles always install the most recent available software
3. These Dockerfiles can almost always build relatively quickly

@@

1. These Dockerfiles require occassional minor maintenance to ensure successful builds. (e.g. changing an installation directive from `unstable` to `testing` or vice versa)
2. The resulting images are inherently dynamic: rebuilding the same Dockerfile months or years apart will generate images with significantly different versions of software installed.
3. The use of apt pinning may be unfamiliar to some users, where the user must ultimately be responsible to ensure compatibility [@apt-pinning].
4. `testing` and `unstable` do not recieve the attention to security updates from the Debian security team.
<!-- DE: Are you sure about that point 4 ? -->

@@

The `debian:testing`-based stack currently includes seven images actively
maintained by the Rocker development team (Table 1).  `r-base` builds on
`debian:testing`, and the other six in the stack each build directly from
`r-base`.  The `r-base` image is unique in that it is designated as the
official image for the R language by the Docker organization itself.
This image is reviewed and built by employees of Docker Inc from the
Dockerfile maintained by the Rocker team.  Consequently, users should
refer to this image in Docker commands without an organization napespace,
e.g. `docker pull r-base` to access the official image.  All other images
in the Rocker project are not individually reviewed and built by Docker
Inc and must be refered to require the organization namespace on Docker
Hub, e.g. `docker pull rocker/r-devel`.

Several of the images in this stack are oriented towards the R development
community. `r-devel`, `drd`, `r-devel-san` & `r-devel-ubsan-clang` all
add a copy of the development version of R side-by-side the current
release of R provided by `r-base`.  On these images, the development
version is aliased to `RD` to distinguish from the current release, `R`.
As the names suggest, each provide slightly different configurations.
Of particular interest are the images providing development R built with
support for C/C++ address sanitizers, which are notoriously difficult
to configure [@edd_sanitizers].

@@

**Build schedule**: The official `r-base` image is rebuilt by Docker
following any updates to the official `debian` images (roughly every
few weeks).  The rest of the stack uses build triggers that rebuild the
images either whenever `r-base` is updated, or the Dockerfile sources are
updated on the corresponding GitHub repository.  The only exception in
this stack is the `drd` image, which is rebuilt each week by cron trigger.

@@

Table: the `debian:testing` image stack

@@

This stack emphasizes stability and reproducibility of the Docker build.
This stack was introduced much more recently (November 2016) 
in response to considerable user input and requests. The key
feature of this stack is the ability to run older versions of R along
with the contemporaneous versions of R packages.  A user specifies the
version desired using an image tag, e.g.  `rocker/r-ver:3.3.1` will
refer to an image with R version 3.3.1 installed.  Omitting the tag
is equivalent to using the tag `latest`, which, as the name implies,
will always point to an image using the current R release.  Thus, users
wanting to create downstream Dockerfiles which are based on the current
release at time but will continue to reconstruct the same environment in
the future after newer R versions are released should explicilty include
the corresponding version tag, e.g. `rocker/r-ver:3.4.0` at the time
of writing, and not the `latest` tag.  Users can also run the current
development version of R using the tag `devel`, which is built nightly
from R-devel sources from `subversion`.

**MRAN archives**: To facilitate installation of only contemporaneous
versions of R packages on these images, the default CRAN mirror from which
to install R packages is fixed to a snapshot of CRAN corresponding to
the last date for which that version of R was the latest release. These
snapshots are provided by the MRAN archive created by Revolution
Analytics (now part of Microsoft). It archives daily snapshots
of all of CRAN from which a user can install packages with the usual
`install.packages()` function  [@MRAN]. Users can always override this
default by passing any current CRAN repository explicitly.  Unlike CRAN,
Bioconductor only updates its repositories through bi-annual releases
aligned to R's Spring release schedule. Thus, Bioconductor packages
can be installed in the usual way using `bioclite`, which automatically
selects the Bioconductor release corresponding to the version of R in use.

**Version tags**:  These version tags are propegated throughout this
stack: for instance, `rocker/tidyverse:devel` will provide the currently
released versions of the R packages in the `tidyverse` [@tidyverse]
installed on the nightly build of R-devel.  Developers building packages
on this stack are encouraged to tag their images accordingly as well.
Table 3 indicates which versions of R are currently available in the
stack, going back to `3.1.0`.  While older versions may be added to
the stack at a later date, we note that the MRAN snapshots began in
starting 2014-09-17 and thus go back only to R `3.1` era.  Each tag must
be built from a separate Dockerfile, enabling minor differences in the
build instructions to accomidate changing dependencies.  Dockerfiles for
past versions (e.g. < `3.4.0` currently) are intended to remain static
over the long term, while the tag for the current version, `latest`,
and `devel` may be tweaked to accomodate new features or dependencies.

@@

The `tidyverse` image contains all required and suggested dependencies
of the commonly used `tidyverse` and `devtools` R packages, including
external database libraries (e.g. MariaDB and PostgreSQL).  Users should
consult the package Dockerfiles or `installed.packages()` list directly
for a complete list of installed packages.  The `verse` library adds
commonly used dependencies, notably a large but not comprehensive LaTeX
environment and Java development libraries.  Previously the Rocker
project provided the image `hadleyverse` which was since divided into
`tidyverse` and `verse` through community input.

@@

- Portable: from laptop to cloud
- Sandboxed: experiment without touching local system
- Optimized: community-tweaked configurations
- Interfaces: Console-based (R, bash, etc) or RStudio\textsuperscript{\textregistered} Server, Shiny\textsuperscript{\textregistered} server
- Extensible: interactively and persistent community extensions
- Versioned:

research use, instructional use

@@

One common use case for Rocker containers is to provide a fast and
reliable mechanism to deploy a custom R envionment to a remote server, such
as Amazon Web Services Elastic Compute (AWS EC2), DigitalOcean, NSF's
Jetstream servers, or private or institutional server hardware.  Being
able to test code in a predictable, pre-configured R environment on a
local machine and then run the same code in an identical environment on a
remote server (e.g. for access to greater RAM, more processors, or merely
to free up the local machine from a long-running computation) is essential
for low-friction scaling of analysis.  Without such contanerization,
getting code to run appropriately in a remote environment can be a major
undertaking, requiring both time and knowledge many would-be users may
not have.

@@

**Root vs user instances**: By default, all rocker containers run as
root. This is consistent with standard Docker practice, allowing both
interactive users and downstream Dockerfiles to install software directly
without having to switch to root.  The main snag of this approach comes if
a user links a local volume and modifies files on that volume, which will
result in the local file being owned by `root` and not the default user.
To avoid this, specify the non-root user at runtime (username `docker`
on the `r-base` stack, or `rstudio` on images deriving from rstudio)
when running an interactive shell.  For users accessing R through RStudio,
this is not necessary.  The docker container will run as root, but a user
logs in through the RStudio server web interface as the non-root user
"rstudio", and thus any changes made to linked volumes will not alter file
permissions on the home directory.  For more details on user permissions,
see the Rocker wiki.

@@

Installation of certain R packages can be challenging,
particularly when those packages require specific external
dependencies. By running R through a Rocker image, users can deploy an
R environment where their desired packages and libraries have already
been installed and configured. As we discuss below, this configuration
benefits from the input of a large community of users, helping to
ensure better and more reliable performance. For instance, one example
of such a configuration is in building and testing R packages using
compiler sanatizers \citep{edd_sanitizers}, which requires R itself
be re-compiled with specialized settings not avaialable or familiar
to most R users.  Using the Rocker images `rocker/r-devel-san` and
`rocker/r-devel-ubsan-clang` provides a fast and simple way for users
to test.

>>>>

All Rocker images are based on the Debian Linux distribution.
The Debian platform provides a small base image, the well-known `apt`
package management system and rich ecosystem of software libraries,
making it the base image of choice for Docker images, including many
of the "official" images maintained by Docker's own development team.
The Debian platform is also perhaps the best supported Linux platform
within the R community, including an active `r-sig-debian` listserve.
The relatively long period between stable Debian releases
(roughly two years recently) means that software in the `debian:stable`
releases can lag significantly behind current releases of popular
software, including R.  More recent versions of packages can be found in
the pre-release distribution, `debian:testing`, while the very latest
binary builds can be found on `debian:unstable`.   The Rocker Project
can be largely divided into two stacks which address different needs,
defined by which Debian distribution they are based on.  The first stack
is based on `debian:testing`, the second, more recently introduced stack
is based only on `debian` stable releases.  These stacks have different
aims and thus provide different images, as shown in Tables 1 & 2.

@@

The `debian:testing` stack aims to makes the most efficient use of
upstream builds -- the precompiled `.deb` binaries provided by the
Debian repositories.  It is both quicker and easier to install software
from binaries, since the package manager (`apt`) manages the necessary
(binary) dependencies and bypasses the time-consuming process of compiling
from source.  Basing this stack on `debian:testing` means that much more
recent versions of commonly-used libraries and compilers are available
as binaries than would be found in a Debian stable release.  In order
to provide access to the most recent avialble binaries, this stack uses
apt-pinning \citep{apt-pinning} to allow the `apt` package manager to also
install binaries from `debian:unstable`, which represents the most recent,
bleeding edge of packages built for Debian when necessary.  For instance,
the `r-base` image provided by Rocker installs the most recent version of
R as a binary from Debian `unstable`.  Similarly, recent versions of many
popular R packages can also be installed through the package manager,
e.g. `apt-get install r-cran-xml`.  This can be particularly helpful
for packages with external system dependencies (such as `libxml2-dev`
in this example) which cannot be installed from the R console as they 
are system dependencies rather than R packages installed from within R.
We should note, however, that only about 500 of the currently almost
11000 CRAN packages are available as Debian packages.

As the names `testing` and `unstable` imply, this approach is not
without challenges.  The particular version of any given package can
change as packages move from `unstable` into `testing`. New versions
are sent to `unstable` during the normal course of Debian development.
This can occassionally break an previously working installation command
in a Dockerfile until the maintainer redirects the package manager to
install a the package from the `unstable` sources that could previously
be installed from `testing`, or vice versa (using the `-t` option in
`apt`, see Box 1 on apt-pinning.)  That said, packages only migrate from 
`unstable` to `testing` after a period of several days---and if the 
migration and installation of the particular version is free of 
interactions with other packages in their dependency graph. That way,
`unstable` serves as validation lab which leaves `testing` reasonably 
stable yet current.

@@

1. These Dockerfiles are easy to develop and extend because almost all software can be installed through the package manager
<!-- DE 1.) would be true of stable too, no?  -->
2. These Dockerfiles always install the most recent available software
3. These Dockerfiles can almost always build relatively quickly

@@

1. These Dockerfiles require occassional minor maintenance to ensure successful builds. (e.g. changing an installation directive from `unstable` to `testing` or vice versa)
2. The resulting images are inherently dynamic: rebuilding the same Dockerfile months or years apart will generate images with significantly different versions of software installed.
3. The use of apt pinning may be unfamiliar to some users, where the user must ultimately be responsible to ensure compatibility \citep{apt-pinning}.
4. `testing` and `unstable` do not recieve the attention to security updates from the Debian security team.
<!-- DE: Are you sure about that point 4 ? -->

@@

The `debian:testing`-based stack currently includes seven images
actively maintained by the Rocker development team (Table 1).
`r-base` builds on `debian:testing`, and the other six in the stack
each build directly from `r-base`.  The `r-base` image is unique in
that it is designated as the official image for the R language by the
Docker organization itself.  This official image is reviewed and then
built by employees of Docker Inc based on the Dockerfile maintained by
the Rocker team.  Consequently, users should refer to this image in
Docker commands without an organization napespace, e.g. `docker pull
r-base` to access the official image.  All other images in the Rocker
project are not individually reviewed and built by Docker Inc and must
be refered to require the organization namespace on Docker Hub,
e.g. `docker pull rocker/r-devel`.

Several of the images in this stack are oriented towards the R
development community. `r-devel`, `drd`, `r-devel-san` &
`r-devel-ubsan-clang` all add a copy of the development version of R
side-by-side the current release of R provided by `r-base`.  On these
images, the development version is aliased to `RD` to distinguish from
the current release, `R`.  As the names suggest, each provide slightly
different configurations.  Of particular interest are the images
providing development R built with support for C/C++ address and
undefined-behavior sanitizers, which are somewhat more difficult to
configure \citep{nedd_sanitizers}.

@@

**Build schedule**: The official `r-base` image is rebuilt by Docker
following any updates to the official `debian` images (roughly every
few weeks).  The rest of the stack uses build triggers that rebuild the
images either whenever `r-base` is updated, or the Dockerfile sources are
updated on the corresponding GitHub repository.  The only exception in
this stack is the `drd` image, which is rebuilt each week by a `cron` trigger.

@@

Table: The `debian:testing` image stack

@@

This stack emphasizes stability and reproducibility of the Docker build.
This stack was introduced much more recently (November 2016) 
in response to considerable user input and requests. The key
feature of this stack is the ability to run older versions of R along
with the contemporaneous versions of R packages.  A user specifies the
version desired using an image tag, e.g.  `rocker/r-ver:3.3.1` will
refer to an image with R version 3.3.1 installed.  Omitting the tag
is equivalent to using the tag `latest`, which, as the name implies,
will always point to an image using the current R release.  Thus, users
wanting to create downstream Dockerfiles which are based on the current
release at time (but will continue to reconstruct the same environment in
the future after newer R versions are released) should explicilty include
the corresponding version tag, e.g. `rocker/r-ver:3.4.0` at the time
of writing, and not the `latest` tag.  Users can also run the current
development version of R using the tag `devel`, which is built nightly
from R-devel sources from `subversion`.

**MRAN archives**: To facilitate installation of only contemporaneous
versions of R packages on these images, the default CRAN mirror from which
to install R packages is fixed to a snapshot of CRAN corresponding to
the last date for which that version of R was the latest release. These
snapshots are provided by the MRAN archive created by Revolution
Analytics (now part of Microsoft). It archives daily snapshots
of all of CRAN from which a user can install packages with the usual
`install.packages()` function  [@MRAN]. Users can always override this
default by passing any current CRAN repository explicitly.  Unlike CRAN,
Bioconductor only updates its repositories through bi-annual releases
aligned to R's spring release schedule. Thus, Bioconductor packages
can be installed in the usual way using `bioclite`, which automatically
selects the Bioconductor release corresponding to the version of R in use.

**Version tags**:  The version tags are propegated throughout this
stack: _e.g._, `rocker/tidyverse:devel` will provide the currently
released versions of the R packages in the `tidyverse` [@tidyverse]
installed on the nightly build of R-devel.  Developers building packages
on this stack are encouraged to tag their images accordingly as well.
Table 3 indicates which versions of R are currently available in the
stack, going back to `3.1.0`.  While older versions may be added to
the stack at a later date, we note that the MRAN snapshots began in
starting 2014-09-17 and thus go back only to the R `3.1` era.  Each tag must
be built from a separate Dockerfile, enabling minor differences in the
build instructions to accomodate changing dependencies.  Dockerfiles for
past versions (e.g. < `3.4.0` currently) are intended to remain static
over the long term, while the tag for the current version, `latest`,
and `devel` may be tweaked to accomodate new features or dependencies.

@@

The `tidyverse` image contains all required and suggested dependencies
of the commonly-used `tidyverse` and `devtools` R packages, including
external database libraries (e.g. MariaDB and PostgreSQL).  Users should
consult the package Dockerfiles or `installed.packages()` list directly
for a complete list of installed packages.  The `verse` library adds
commonly-used dependencies, notably a large but not comprehensive LaTeX
environment and Java development libraries.  Previously the Rocker
project provided the image `hadleyverse` which was since divided into
`tidyverse` and `verse` through community input.

@@

- Portable: from laptop to cloud
- Sandboxed: experiment without touching local system
- Optimized: community-tweaked configurations
- Interfaces: Console-based (R, bash, etc) or RStudio\textsuperscript{\textregistered} Server, Shiny\textsuperscript{\textregistered} server
- Extensible: interactively and persistent community extensions
- Versioned: research use, instructional use

@@

One common use case for Rocker containers is to provide a fast and
reliable mechanism to deploy a custom R envionment to a remote server, such
as Amazon Web Services Elastic Compute (AWS EC2), DigitalOcean, NSF's
Jetstream servers, or private or institutional server hardware.  Being
able to test code in a predictable, pre-configured R environment on a
local machine and to then run the same code in an identical environment on a
remote server (e.g. for access to greater RAM, more processors, or merely
to free up the local machine from a long-running computation) is essential
for low-friction scaling of analysis.  Without such contanerization,
getting code to run appropriately in a remote environment can be a major
undertaking, requiring both time and knowledge many would-be users may
not have.

@@

**Root vs user instances**: By default, all rocker containers run as
root. This is consistent with standard Docker practice, allowing both
interactive users and downstream Dockerfiles to install software directly
without having to switch to root.  The main snag of this approach comes if
a user links a local volume and modifies files on that volume, which will
result in the local file being owned by `root` and not the default user.
To avoid this, specify the non-root user at runtime (username `docker`
on the `r-base` stack, or `rstudio` on images deriving from rstudio)
when running an interactive shell.  For users accessing R through RStudio,
this is not necessary.  The docker container will run as root, but a user
logs in through the RStudio server web interface as the non-root user
"rstudio", and thus any changes made to linked volumes will not alter file
permissions on the home directory.  For more details on user permissions,
see the Rocker wiki. [@rockerwiki]

@@

Installation of certain R packages can be challenging,
particularly when those packages require specific external
dependencies. By running R through a Rocker image, users can deploy an
R environment where their desired packages and libraries have already
been installed and configured. As we discuss below, this configuration
benefits from the input of a large community of users, helping to
ensure better and more reliable performance. For instance, one example
of such a configuration is in building and testing R packages using
compiler sanitizers \citep{edd_sanitizers}, which requires R itself
be re-compiled with specialized settings not avaialable or familiar
to most R users.  Using the Rocker images `rocker/r-devel-san` and
`rocker/r-devel-ubsan-clang` provides a fast and simple way for users
to test.
